Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-03-31 ms.tex
the ghost of balmaverde edited 6.3

It is also possible that many of the galactic nuclei from the
\citet{Miller+15} sample are not accreting in a steady fashion.  The
efficiency assumed in the right panel of Figure \ref{fig:miller} in
some ways represents a maximum value, since it does not take into
suppression of the accretion rate due to disk outflows on small
scales.  If the actual X-ray efficiency is lower than assumed, then
many of the detections in the case of low SMBH masses, $M_{\bullet}
\lesssim 10^{8}M_{\odot}$, would lie even further above above the
predicted average value for stable accretion (black line).  These
galaxies either experience significantly lower levels of heating than
average galaxies of their mass, or that they are not accreting stably
at all.  Indeed, for the lowest mass holes, $M_{\bullet} \lesssim
10^{8}M_{\odot}$, $L_{\rm X}$ nearly exceeds that predicted for the
maximum thermally-stable accretion rate, $\dot{M}_{\rm TI}$ (yellow
line).  It is not surprising that this sample is more likely to be
thermally unstable as compared to an average galaxy with this black
hole mass, because the \citet{Miller+15} sample includes only
early-type galaxies with on average older stellar populations.

Figure \ref{fig:bh_growth} (top panel) also shows that across all
black hole masses, Compton heating also cannot prevent average
accretion rates which are thermally unstable, i.e. $\dot{M}_{\rm C} >
\dot{M}_{\rm TI}$.  In the case of thermal instabilities, nuclear gas
may undergo cyclic oscillations of rapid cooling and high accretion
rates, followed by quiescent periods once gas has been consumed and/or
AGN feedback becomes effective.  Such cycles are proposed to occur on
large scales and long timescales in elliptical galaxies
(\citealt{Ciotti+10}), an idea possibly supported by, e.g., GALEX
observations showing that a small fraction of early-type galaxies in
the local universe have undergone recent ($< 1-2$ Gyr old) star
formation episodes (\citealt{Donas+07}).  Similar cyclic AGN activity
have been proposed to occur on the smaller scale of the sphere of
influence (e.g.~\citealt{Yuan&Li11}, \citealt{Cuadra+15}).

If such cyclic outbursts also manifest on the properties of gas on
small, parsec scales, then the large scatter in the accretion rates
and hence X-ray luminosities in Fig \ref{fig:miller} could result from
the wide range of inflow rates experienced over the course of a cyclic
episode between instabilities and periodic nuclear outbursts
(e.g.~\citealt{Ciotti+10}).  Due to their deeper potential well, we
find that core galaxies are more likely to possess higher accretion
rates and potentially be thermally unstable for a fixed wind heating
rate than cusp galaxies (Fig.~XXX).  This could help explain why core
galaxies show a wide range of nuclear radio and X-ray luminosities
(suggesting cyclic accretion episodes), while cusp galaxies are
confined below a threshold (e.g., \citealt{Bender+89, Pellegrini99,
  Capetti&Balmaverde05}), consistent with thermally-stable,
steady-state picture. ({\bf AG: This picture may be dubious...Review
  the literature but I think that the higher upper bound for cores is
  just due to a mass selection effect.})

If the galaxies with measured $L_{\rm X}$ at low $M_{\bullet}$ in Fig
\ref{fig:miller} represent thermally-unstable galaxies, then many of
the upper limits may correspond to a separate ``thermally stable"
population, for instance including SgrA*.  The possibility of a mixed
population of stable versus unstable galaxies calls into question the
use of simple extrapolations of power-law fits to the $L_{\rm
  X}(M_{\bullet}$) relationship to low $M_{\bullet}$ to constrain the
occupation fraction of black holes in low mass galaxies
(\citealt{Miller+15}).

>>>>

It is also possible that many of the
\citet{Miller+15} sample of galactic nuclei are not accreting in steady state.  The
value of the X-ray efficiency assumed in the right panel of Figure \ref{fig:miller} represents a maximum value, since it does not account for suppression of the SMBH accretion rate due to outflows from the accretion disk on small
radial scales.  If the actual radiative efficiency is lower, then many of the X-ray detections would lie even further above above the predicted value of $\langle L_X \rangle$ for stable accretion (black line).  Either these
galaxies experience much less heating than average galaxies of the same mass, or they are not accreting stably
at all.  Indeed, for $M_{\bullet} \lesssim
10^{8}M_{\odot}$, $L_{\rm X}$ nearly exceeds that predicted for the
maximum thermally-stable accretion rate, $\dot{M}_{\rm TI}$ (yellow
line).  Since the \citet{Miller+15} sample includes only early-type galaxies with older stellar populations than the average for their black hole mass, it would not be too surprising if many of these galaxies were thermally unstable.

Figure \ref{fig:bh_growth} (top panel) also shows that Compton heating cannot prevent average
accretion rates which are thermally unstable, i.e. $\dot{M}_{\rm C} >
\dot{M}_{\rm TI}$, across all
black hole masses.  When thermal instabilities become manifest, nuclear gas may undergo cyclic oscillations of rapid cooling and high accretion rates, followed by quiescent periods once gas has been consumed and/or
AGN feedback becomes effective.  Such cycles are seen in numerical simulations of elliptical galaxies on 
large scales and long timescales (\citealt{Ciotti+10}), an idea which supported by observations showing that a fraction of early-type galaxies in the local universe have undergone recent ($< 1-2$ Gyr old) star formation episodes (\citealt{Donas+07}).

Similar cyclic AGN activity could occur on the smaller radial scale of the sphere of influence (e.g.~\citealt{Yuan&Li11}, \citealt{Cuadra+15}).  In this case the large scatter in the X-ray luminosities shown in Fig \ref{fig:miller} could result from
the wide range of inflow rates experienced over the course of a cyclic episode between instabilities and periodic nuclear outbursts.  Due to their deeper potential wells, we find that core galaxies are more likely to possess higher accretion
rates and potentially be thermally unstable for a fixed wind heating rate than cusp galaxies.  This could help explain why core galaxies show a wide range of nuclear radio and X-ray luminosities (suggesting cyclic accretion episodes), while cusp galaxies are confined below a threshold (e.g., \citealt{Bender+89, Pellegrini99,   Capetti&Balmaverde05}), consistent with thermally-stable, steady-state picture. ({\bf AG: This picture may be dubious...Review
  the literature but I think that the higher upper bound for cores is
  just due to a mass selection effect.  BDM: The ghost of Balmaverde disagrees.})

If the galaxies for which $L_{\rm X}$ is measured at low $M_{\bullet}$ in Fig
\ref{fig:miller} indeed represent thermally-unstable galaxies, then a fraction of the upper limits may contribute to a separate thermally stable population, for instance including SgrA*.  However, the possibility of a mixed population of stable versus unstable galaxies calls into question the use of simple extrapolations of power-law fits to the $L_{\rm X}(M_{\bullet}$) relationship to low $M_{\bullet}$ to constrain the occupation fraction of black holes in low mass galaxies (\citealt{Miller+15}).
